---
title: Did joining the EC/EU benefit the UK economy?
author: Bob Beachill
date: '2019-04-22'
slug: did-joining-the-ec-eu-benefit-the-uk-economy
categories:
  - UK Economy
  - EU
tags:
  - GDP
  - growth
  - UK
  - EU
description: ''
thumbnail: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

```{r include=FALSE}
library(forecast)
library(ggplot2)
library(readxl)
```

Given all the discussion about whether leaving the European Union will cause long run harm to the UK economy, it is instructive to look at the impact that joining the EU (or European Community (EC) as it then was) back in 1973 had on the UK’s long run economic performance.

## How to assess the impact?
Obviously to carry out a detailed assessment and evaluation of the impact that joining the EC/EU had on the UK economy would involve a huge amount research and analysis looking at the effect on various sectors and industries. All I intend to do here is simply to look at the growth in average living standards, measured by real GDP per head, in the UK before and after the UK’s accession. But just comparing growth before and after accession isn’t sufficient as we don’t know how the UK economy would have performed if the UK had not joined. Just because, say, growth improved after 1973 doesn’t necessarily mean joining the EC/EU was beneficial as growth might have improved anyway, and maybe even more so. Plus there will be other factors at work affecting growth after 1973 (and before). One way to approach this issue, which I think is valid, is to to look at *comparative growth*, i.e. compare the UK’s growth record with that of other similar countries before and after its accession. These countries, specifically Germany, France and Italy are all in the EC/EU throughout the entire period. Thus, if the factors which affect growth are experienced by all the countries being compared then we have a kind of controlled, natural experiment where the only significant difference between them in the before and after periods is UK membership. It is therefore possible, albeit not certain, to ascribe at least some of the change in the UK’s growth record after accession to its membership.

## UK growth before joining in 1973
So what does the data show? Figure 1 compares growth in annual real GDP per head of West Germany (as it then was), France, Italy and, of course, the UK. The source this data is the [Annual Macroeconomic Database of the European Commission (AMECO)](http://ec.europa.eu/economy_finance/ameco/user/serie/SelectSerie.cfm){target=“_blank”}. The data is GDP per head of population at constant prices (code RVGDP). I have indexed the data to a base of 1960 (the earliest date available) and shown the time path up to 1973 which is the year of the UK’s accession to the EC.

```{r include=FALSE}
uk_euro <-
  read_excel("~/OneDrive/Documents/Macro Data/Eurostat Data/uk_euro.xlsx",
             range = "A1:D15")
uk_euro_before <- ts(uk_euro, start = 1960)
```


```{r echo=FALSE, fig.cap="Real GDP per head, 1960-73"}
autoplot(uk_euro_before, lwd=1.1) + ggtitle("Real GDP per head, annual CVM, 1960-1973\n(1960=100)") + ylab("Index") + xlab("") + theme(legend.position = "bottom") + guides(colour=guide_legend(title=NULL)) + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + annotate("label", x = 1963, y = 160, label = "Source: AMECO, Code RVGDP")
```

What can easily be seen is that during this period the UK economy suffered economic decline relative to the other three major European economies. The UK’s real GDP per head grew by 45% (annual average growth rate of 2.89%) from 1960 through to 1973, whereas for West Germany the growth rates were 56% (3.46%), for Italy 81% (4.65%), and for France 79% (4.58%). In consequence, the gap narrowed between the levels of real GDP per head of these three countries and the UK. It did not necessarily mean that by 1973 their average standard of living had overtaken the UK, given that they all started from different levels. In fact, as Figure 3 below shows, by 1973 only Italy’s real GDP per head was still below that of the UK (and only marginally).

Can we blame this relative decline of the UK economy on its non-membership of the EC unlike the other countries? With certainty, definitely not. But this evidence is surely suggestive, and it would be difficult to contend that it was not an important factor. In fact it can be [argued](https://voxeu.org/article/britain-s-eu-membership-new-insight-economic-history){target=“_blank”} it was the desire to stop and, hopefully, reverse this comparative decline that was the principal reason behind the determination of the Heath-led conservative government to join the EC.^[I’m old enough to have been around at the time and remember this being one of the key arguments in favour of seeking membership]

## UK growth after joining
Let’s look now at the corresponding data for the period after the UK’s accession. Figure 2 shows the time path of real GDP per capita for the period 1974 to 2018. Again the data has been indexed to a base of 1974 for each country.

```{r include=FALSE}
uk_euro1 <- read_excel("~/OneDrive/Documents/Macro Data/Eurostat Data/uk_euro.xlsx", 
                      range = "F1:I46")
uk_euro_after <- ts(uk_euro1, start = 1974)
```

```{r echo=FALSE, fig.cap="Real GDP per head, 1974-2018"}
autoplot(uk_euro_after, lwd=1.1) + ggtitle("Real GDP per head, annual CVM, 1974-2018\n(1974=100)") + ylab("Index") + xlab("") + theme(legend.position = "bottom") + guides(colour=guide_legend(title=NULL)) + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + annotate("label", x = 1985, y = 200, label = "Source: AMECO, Code RVGDP")+ geom_vline(xintercept=1999, colour = "blue") + annotate("label", x = 1999, y = 110, label = "Euro")
```

The contrast with the pre-accession period is striking and significant. Since joining, the UK economy has grown in line with the other three countries and indeed it has outpaced them since the late 1990s, especially since the introduction of the monetary union in 1999.^[Most macroeconomists would agree that the UK made the correct decision to not adopt the Euro as its currency.] Over the entire 45 year period the UK’s average standard of living grew by 117% (annual average growth rate of 1.78%). The figures for Germany^[West Germany until 1990] are 89% (1.46%), for Italy 72% (1.24%), and France 84% (1.39%). So since joining the UK has managed to reverse some of the relative economic decline it experienced before 1973. Again, we cannot be certain that this improvement in its relative performance is solely or even primarily down to it “joining the club”; perhaps it would have happened anyway, or have been even better if the UK had remained outside the EC/EU. But given the significant difference between pre and post accession relative growth it would very difficult to argue, I think, that the UK’s membership was not a major, if not the major, cause of this difference.

Two other points of note are:

* The impact on Germany’s real GDP per head of reunification in 1990 is clear to see
* The impact of the Great Recession of 2008-09 on the growth of average living standards is also very apparent

## Average living standards before and after
A final way to see the overall impact on the UK is to compare the average living standard in the UK to the other three countries examined here both before its accession to the EC/EU and today. Figure 3 shows the ratio of each country’s GDP per head (measured in current prices and converted to US dollars using purchasing power parity exchange rates^[PPPs are used to ensure that one dollar buys the same amount of goods in each country. A succinct explantion [here](https://www.economicsonline.co.uk/Global_economics/Purchasing_power_parity.html){target=“_blank”}. More detail [here](https://en.wikipedia.org/wiki/Purchasing_power_parity){target=“_blank”}.]) to the UK in 1972 and 2018. The data is taken from the [OCED.Stat](https://stats.oecd.org){target=“_blank”} site.

In 1972, the UK’s average standard of living was below that of France and West Germany and almost identical to Italy. In 2018, after 45 years of membership, the UK’s GDP per head is now above that of France, albeit marginally, well above Italy, but still lower than Germany.

```{r include=FALSE}
GDP_per_head_PPS <- read_excel("~/OneDrive/Documents/OECD Data/GDP per head PPS.xlsx", 
    range = "L9:N17", col_types = c("text", 
        "text", "numeric"))
```

```{r echo=FALSE, fig.cap="Ratio to UK GDP per head (Source: OECD)"}
ggplot(GDP_per_head_PPS, aes(x=Year, y=RatioUK, fill = Country)) +
  geom_bar(position = "dodge", stat = "identity", colour = "black") + ggtitle("Ratio to UK GDP per head, 1972 & 2018\n(US dollars, current prices, PPPs)") + ylab("Ratio") + xlab("") + guides(colour=guide_legend(title=NULL)) + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + scale_y_continuous(limits = c(0,1.4))
```

