---
title: Some facts (and myths) about government deficits and debt
author: Bob Beachill
date: '2019-06-22'
slug: some-facts-and-myhs-about-government-deficits-and-debt
categories:
  - UK Economy
  - Government
tags:
  - Fiscal Policy
description: ''
thumbnail: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

```{r include=FALSE}
library(forecast)
library(ggplot2)
library(pdfetch)
library(readxl)
```

Ever since the Great Recession, and the policies which many government’s enacted in response, led to a significant rise in their deficits and debt, there has been repeated discussion about the nature and sustainability of public finances. Unfortunately, much of this discussion has been ill-informed  and confused. In this post, I will attempt to give a clear explanation of what the government deficit and debt levels are, how they are connected, and how they evolve over time.

## What is the deficit, what is the debt?
Some people, including journalists and politicians, sometimes confuse the government’s deficit with its debt.

- The government’s overall budget balance *over a given period of time*, say a quarter or a year, is the difference between its total income and its total expenditure. If income exceeds expenditure, it has a surplus; if its expenditure exceeds its income, it has a deficit.
- The level of the government’s debt is the total amount of borrowing outstanding (i.e. money owed) *at a particular point in time*, say, for example, at the end of the financial year.
- If the government has a surplus it will use this to pay off some debt and thus reduce its level; conversely, if there is a deficit, the level of debt will increase.

According to the [UK Office of National Statistics (ONS)](https://www.ons.gov.uk/economy/governmentpublicsectorandtaxes/publicsectorfinance/bulletins/publicsectorfinances/april2019), “Borrowing in the latest full financial year (April 2018 to March 2019) was £23.5 billion.” This was the overall deficit of the UK public sector which includes central government, local government and public corporations. Also, according to the ONS, “Debt (public sector net debt excluding public sector banks) at the end of April 2019 was £1,797.7 billion.”

## Deficit and debt as % of GDP
These sound enormous figures and politicians and other commentators often use them to sound the alarm over the “parlous” state of the public finances. But it is not the absolute amounts of the deficit and debt which really matter. It’s the size of the deficit and debt as a percentage of Gross Domestic Product (GDP) which is important. Imagine two individuals each with a total debt of, say, £2000. The first person has an income of £10000 per year, the second has an annual income of £100000. Although the absolute level of debt is the same for both individuals, it’s clear that the burden of that debt is much higher for the first person than the second. Servicing and repaying a debt which is one fifth of your income is much harder than a debt which is only one fiftieth. So, it’s important to consider the *relative* size of debt (to income) rather than just the absolute amount.

When it comes to the deficit and debt of the government, its income is effectively national income, i.e. GDP, as it is this which constitutes the tax base. The larger is GDP the more taxation revenue the government could potentially raise in order to service and pay off the debt. Hence, the larger is GDP for any given absolute level of debt, the lower the burden of the debt on the public finances^[A similar argument applies to the deficit]. Again, according to the [UK Office of National Statistics (ONS)](https://www.ons.gov.uk/economy/governmentpublicsectorandtaxes/publicsectorfinance/bulletins/publicsectorfinances/april2019), public sector net debt (excluding public sector banks) was 82.7% of GDP at the end of April 2019, whilst public sector borrowing (the deficit) in the latest full financial year (April 2018 to March 2019) was 1.1% of GDP.

## Deficit and Debt over time
Are these UK deficit and debt ratios particularly high compared to the past? Looking at the historical record, the charts below show the deficit (net borrowing) and debt ratios, from 1956 for the deficit and from 1975 for the debt^[Earlier figures for the debt ratio are not available on the ONS site].

```{r echo=FALSE, fig.cap="UK Deficit/GDP ratio", message=FALSE}
UKdefcitgdp <- pdfetch_ONS("J5IJ","PUSF")
UKdefcitgdp <- na.omit(UKdefcitgdp)
UKdefcitgdp <- ts(UKdefcitgdp, start = 1956)
autoplot(UKdefcitgdp) + ggtitle("UK Public Sector Net Borrowing as % of GDP\n(excl public sector banks), 1955-56 to 2018-19") + ylab("Percent") + xlab("") + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + scale_x_continuous(limits=c(1955,2020), breaks=seq(1955,2020,by=5)) + annotate("label", x = 1970, y = 8.0, label = "Source: ONS, code J5IJ")
```

In Figure 1, the deficit to GDP ratio appears to be on a rising trend over the period from financial year 1955-1956 to 2018-2019, but with a lot of fluctuations.These fluctuations around trend correspond, for the most part, to the booms and slumps in the economy. When the economy is booming, i.e. GDP growth is relatively high, the deficit ratio will tend to fall; conversely, when the economy goes into recession or experiences lower than normal growth, the deficit as a ratio of GDP will tend to increase. There are two reasons why the deficit ratio tends to rise, particularly in a recession when GDP growth is negative. Firstly, tax revenue will automatically fall as incomes fall and people spend less, and expenditures on unemployment benefits automatically will increase: this is the impact of the “automatic stabilisers”, so called because they automatically help to stabilise the fall in aggregate spending in the economy. Secondly, the government might also carry out discretionary changes in tax rates and expenditure, specifically, they will cut tax rates and boost spending in order to further support total spending in the economy.

The clearest example of these factors at work is the large rise in the deficit ratio as a result of the onset and policy response to the Great Recession of 2008-09. The ratio in the financial year 2007-08 was a comparatively modest 2.8%, fairly average for the period up until then. But as the recession took hold, the automatic stabilisers in conjunction with the discretionary fiscal stimulus enacted by the UK government caused the ratio to increase quickly and significantly to 7.3% in 2008-09 and then to a high of 9.9% in 2009-10. In response to this significant rise in the deficit ratio the government embarked on its "austerity" program and this, together with a resumption of economic growth, has resulted, as can be seen in the chart, in a sustained reduction in the deficit to GDP ratio.

```{r echo=FALSE, fig.cap="UK Debt/GDP ratio", message=FALSE}
UK_Net_Debt_as_of_GDP <- read_excel("~/OneDrive/Documents/Macro Data/UK Net Debt as & of GDP.xlsx")
UKNetDebt <- ts(UK_Net_Debt_as_of_GDP[,-1], start = 1975)
autoplot(UKNetDebt) + ggtitle("UK Public Sector Net Debt as % of GDP\n(excl public sector banks), end Mar 1975 to end Mar 2019") + ylab("Percent") + xlab("") + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + scale_x_continuous(limits=c(1975,2020), breaks=seq(1975,2020,by=5)) + annotate("label", x = 1985, y = 70.0, label = "Source: ONS, code HF6X")
```

Looking at Figure 2 showing the time path of the debt/GDP ratio, from a comparative high of 55.8% at the end of financial year 1974-75 the ratio gradually declined until the onset of the Great Recession, but again with noticeable fluctuations. At the end of 2007-08 it was 35.2%, but having been as low as 24.2% at financial year end 1990-91. The dramatic rise in the debt ratio as a consequence of the Great Recession and its aftermath is obvious and clear to see. The ratio increased year after year and until it reached a peak of 85.1% at year-end March 2017. This significant rise is of course linked, and indeed a direct consequence of, the rise in the deficit ratio. The link between the deficit and the debt ratios is, however, not one-to-one. Notice that although the deficit/GDP ratio has been reduced considerably from its peak of 9.9% down to 1.1% in financial year ending March 2019, the debt/GDP ratio has remained at its very high level, only falling slightly to 83.1% by the end of March 2019.