---
title: How well is the UK economy doing?
author: Bob Beachill
date: '2019-04-13'
slug: how-well-is-the-uk-economy-doing
categories:
  - UK Economy
tags:
  - GDP
  - growth
  - UK
description: ''
thumbnail: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

A question which crops up repeatedly in political discussions concerns the macroeconomic performance of the UK economy since the ending of the Great Recession of 2008-09. One way to answer this question is to examine the time path of **UK real GDP per head** both historically and in comparison with other similar industrialised nations.

## Real GDP per head
Why real GDP per head? What exactly is that? GDP stands for Gross Domestic Product - it's a measure of the value of the total output of goods and services produced in an economy in a given period. Now in order to calculate the value of output of any firm or industry the government statisticians have to use the current prices of each product or service. This measure is known as GDP in current prices. But these prices change over time. So that presents a problem in determining whether the economy is producing a greater volume of goods and services in one period compared to another. Just comparing GDP in current prices in different periods won't tell us because this measure will change partly because of volume changes but also partly because of changes in prices. So in order to determine whether an economy is producing a greater or lesser volume of goods and services in one period compared to another we need a measure of GDP that removes the effects of changing prices: this is called Real GDP. (Technically, it involves measuring GDP at chained volume measures (CVM) rather than current prices)^[A technical description of how GDP is calculated can be found in the [UK Blue Book](https://www.ons.gov.uk/economy/grossdomesticproductgdp/compendium/unitedkingdomnationalaccountsthebluebook/2018/anintroductiontotheuknationalaccounts){target="_blank"} ]

Many politicians and journalists look at the growth of real GDP as an indicator of macroeconomic performance, and in some cases this is valid. But real GDP will increase if a country experiences growth in its labour force (perhaps as a result of positive net migration), and this may not necessarily mean a rise in GDP per head of population. *It is real GDP per head that measures the average standard of living in a country*, so this is what we really ought to look at when assessing the growth performance of an economy.

## UK Real GDP per head over time
Here is a chart (Figure 1) showing the time path of quarterly UK real GDP per head since 1955Q1. There has been a gradual upward trend as average living standards have increased over time. There have also been periods of recession when growth has been negative, none more so than in the Great Recession of 2008 - 2009. The line is curving upwards indicating an exponential growth trend of some average percentage amount each quarter. To see this more clearly we can employ the standard technique of plotting the natural logarithm of the data. This turns an upward sloping exponential trend into a linear trend where the slope of the line equals the growth rate.^[Here’s the maths$$
Y_t=Y_0e^{rt}\,\,\,\,\,\,\, \text{Exponential growth}\\
logY_t=logY_0+rt\,\,\,\,\,\,\, \text{Taking natural logs}
$$ and here is more [info](https://en.wikipedia.org/wiki/Exponential_growth){target="_blank"} if you want it.]

```{r include=FALSE}
library(pdfetch)
library(forecast)
library(ggplot2)
library(dyn)
```

```{r echo=FALSE, fig.cap= "UK Real GDP per head"}
data2 <- pdfetch_ONS("IHXW", "UKEA")
data2 <- ts(data2, start = c(1955,1), frequency = 4)

autoplot(data2) + ggtitle("UK GDP (average) per head, CVM market prices, SA\n1955Q1 - 2018Q4") + ylab("£, CVM") + xlab("") + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + annotate("label", x = 2000, y = 2500, label = "Source: ONS, code IHXW")
```

```{r include=FALSE}
lndata2 <- log(data2)
lndata2
trend2 <- time(lndata2)

mod2 <- dyn$lm(window(lndata2, start=c(1955,1), end=c(2007,4)) ~ trend2)
summary(mod2)
newdata2 <- window(trend2, start=c(2008,1))
extrap2 <- predict.dyn(mod2,newdata = newdata2)

gdpcap <- cbind(lndata2,extrap2)
colnames(gdpcap) <- c("Actual","Trend: 1955Q1-2007Q4")
```

```{r echo=FALSE, fig.cap= "UK Log Real GDP per head"}
autoplot(gdpcap) + ggtitle("UK Log GDP (average) per head, CVM market prices, SA\n1955Q1 - 2018Q4") + ylab("log") + xlab("") + theme(legend.position = "bottom") + guides(colour=guide_legend(title=NULL)) + scale_x_continuous(limits=c(1950,2020), breaks=seq(1950,2020,by=10)) + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + annotate("label", x = 2000, y = 8.0, label = "Trend growth = 0.58 % per quarter")
```

Figure 2 shows the time path of the log of real GDP per head. The exponential tend has been straightened out into a linear trend. What’s now noticeable is that the trend growth rate from 1955Q1 to 2007Q4 (just before the start of the Great Recession) is higher than the growth rate since the end of the recession in 2009Q3. To find the average trend growth rate for the period 1955Q1 to 2007Q4 a line of best fit can be calculated and superimposed on the chart as shown. So between 1955Q1 and 2007Q4, real GDP per head in the UK grew on average by 0.58% per quarter (i.e. about 2.32% per year). There were periods when growth was higher than that (“booms”) and periods when it was lower (“recessions”). But over this period the economy tended to always get back to trend fairly quickly following these episodes of higher or lower growth. In particular, following a recession the economy *grew faster than trend* in order to get back to where it would have been had it just continued growing steadily at the average rate.

The trend line has been extended to show where real GDP per head would have grown to if, following the end of the Great Recession, the UK economy economy had once again returned to this average growth path. It’s obvious that this time that has not happened. Rather than growing faster in order to catch back up to this long run trend real GDP per head has grown more slowly than the previous average such that the gap between actual and trend is large and widening. The average growth rate since 2009Q3 has been 0.3% per quarter (i.e. about 1.2% per year). In consequence, real GDP per head in 2018Q4 is 19.5% below the previous trend level. Put the other way, real GDP per head could be 24% higher than it is in 2008Q4 if the economy had returned to its previous trend. In money terms, the average person is £1861 per quarter worse off.

From an historical viewpoint it’s fair to conclude that the performance of the UK economy since the onset and end of the Great Recession has been very poor indeed. Average growth per quarter in real GDP per head has been half the previous long run rate. Hence, unlike following previous recessions, the economy has failed to get back to its previous trend level. There are a number of possible explanations for this weak performance but I will discuss these in a future post.

The other question to consider is whether this reduced performance is unique to the UK. Have other advanced industrial nations also grown much more slowly since the Great Recession than in the past? And if so has the UK’s performance been better or worse?

## UK compared to other countries
The UK is not unique in experiencing much lower growth in average living standards since the end of the Great Recession. In fact, most other advanced industrial nations including the USA, Germany, France and Italy have suffered similarly, and for the same reasons (almost certainly) as the UK. Take for example the US. Figure 3 shows the time path of log real GDP per head with the trend growth line for 1955Q1 to 2007Q4 superimposed. As can be seen the picture is very similar to that of the UK.

```{r include=FALSE}
USgdpcap <- pdfetch_FRED("A939RX0Q048SBEA")
USgdpcap <- ts(USgdpcap, start = c(1947,1), frequency = 4)
USgdpcap
autoplot(USgdpcap)
USgdpcap_55 <- window(USgdpcap, start = c(1955,1))
USgdpcap_55
autoplot(USgdpcap_55)
lndata3 <- log(USgdpcap_55)
lndata3
trend3 <- time(lndata3)

mod3 <- dyn$lm(window(lndata3, start=c(1955,1), end=c(2007,4)) ~ trend3)
summary(mod3)
newdata3 <- window(trend3, start=c(2008,1))
extrap3 <- predict.dyn(mod3,newdata = newdata3)

gdpcap3 <- cbind(lndata3,extrap3)
colnames(gdpcap3) <- c("Actual","Trend: 1955Q1-2007Q4")
```

```{r echo=FALSE, fig.cap= "US Log Real GDP per head"}
autoplot(gdpcap3) + ggtitle("US Log GDP per head, Chained 2012 dollars, SAAR\n1955Q1 - 2018Q4") + ylab("log") + xlab("") + theme(legend.position = "bottom") + guides(colour=guide_legend(title=NULL)) + scale_x_continuous(limits=c(1950,2020), breaks=seq(1950,2020,by=10)) + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + annotate("label", x = 2000, y = 10, label = "Source: FRED")
```

So, given that this phenomenon is not exclusive to the UK, how does the UK’s performance compare. Figure 4 shows how annual real GDP per head has grown since 2009. It compares the UK with a range of other advanced economies plus the Euro area (12 countries).^[Austria, Belgium, Finland, France, Germany, Ireland, Italy, Luxembourg, Netherlands, Portugal, Spain, Greece] Compared to France, Italy (especially) and the Euro 12 area as a whole the UK has done reasonably well. However, its performance has lagged behind Germany and the US. Whether or not the UK economy should have performed as weals those two latter countries is open to debate.

```{r include=FALSE}
library(readxl)
Real_GDP_per_head_2009_2018 <- read_excel("~/OneDrive/Documents/Macro Data/Eurostat Data/Real GDP per head 2009-2018.xlsx")
rgdphead <- ts(Real_GDP_per_head_2009_2018[,-1], start = 2009)
```

```{r echo=FALSE, fig.cap= "Real GDP per head, various countries"}
autoplot(rgdphead, lwd=1.2) + ggtitle("Real GDP per head, annual CVM, 2009-2018\n(2009=100)") + ylab("Index") + xlab("") + theme(legend.position = "bottom") + guides(colour=guide_legend(title=NULL)) + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + annotate("label", x = 2011, y = 115, label = "Source: Eurostat, FRED") + scale_color_brewer(palette = "Set1")
```

