---
title: Test Post of Plotly chart
author: Bob Beachill
date: '2019-05-05'
slug: test-post-of-plotly-chart
categories: []
tags: []
description: ''
thumbnail: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

```{r include=FALSE}
library(ggplot2)
library(readxl)
library(plotly)
```

This is just a test post to see ifthe ggplot charts I create are animated correctly by the plotly package in R. Look slike I'll need alot,ore text here atthe start.

This is just a test post to see ifthe ggplot charts I create are animated correctly by the plotly package in R. Look slike I'll need alot,ore text here atthe start.

This is just a test post to see ifthe ggplot charts I create are animated correctly by the plotly package in R. Look slike I'll need alot,ore text here atthe start.


```{r include=FALSE}
uk_euro <-
  read_excel("~/OneDrive/Documents/Macro Data/Eurostat Data/uk_euro.xlsx",
             range = "A1:D15")
Year <- 1960:1973
uk_euro <- cbind(uk_euro,Year)
```

```{r echo=FALSE}
p <- ggplot(uk_euro, aes(Year,Italy)) + geom_line()
ggplotly(p)
```

