---
title: Austerity; What is it, how much has there been, was it worth it?
author: Bob Beachill
date: '2019-08-17'
slug: austerity-what-is-it-how-much-has-there-been-was-it-worth-it
output:
  blogdown::html_page:
    toc: true
categories:
  - UK Economy
  - Government
tags:
  - Fiscal Policy
  - UK
description: ''
thumbnail: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

```{r include=FALSE}
library(forecast)
library(ggplot2)
library(pdfetch)
library(readxl)
UK_Austerity <- read_excel("~/OneDrive/Documents/Blog Posts data/UK Austerity.xlsx")
colnames(UK_Austerity) <- c("Year","rcexp","rnetinv","cycpbal","cexpgdp","netinvgdp","cycpbalgdp","ukpop")
UKAust <- ts(UK_Austerity[,-1], start = 1955)
UKAust75 <- window(UKAust, start = 1975)
```

One of the most controversial, and hotly debated concepts in economics in recent years is austerity. In this post I will consider what the term “austerity” actually means in economics, and I’ll examine the available evidence on the manner and extent of austerity that has taken place in the UK since 2010. I will also briefly look at austerity in other countries.

## What’s the definition of austerity?
Sometimes the term austerity is used in a general sense to refer to difficult economic conditions when shortages and rationing of goods might be imposed. This was the case in the UK and other countries after the second world war. However, the more recent usage of the term refers to the fiscal consolidations enacted by several countries following the Great Recession. These consolidations involved reductions in public expenditure and/or increases in taxation in order to try to reduce the government’s deficit and debt. But should all fiscal contractions always be regarded as austerity? Not necessarily. As with many macroeconomic policies it depends on the current state of the economy. If an economy is operating at close to full employment or experiencing a boom, some fiscal consolidation might be required to reduce inflationary pressures or to free resources to be used by the private sector. The term austerity is better reserved for a fiscal contraction undertaken when an economy is suffering from a recession or a depression. Such a consolidation is likely to make the recession/depression worse as it reduces aggregate spending in the economy still further^[Unless you believe in the theory of "expansionary austerity"], unless it can be countered by a more expansionary monetary policy (i.e. a cut in interest rates). Simon Wren-Lewis, a well known macroeconomist, [defines austerity](https://mainlymacro.blogspot.com/2017/09/defining-austerity-redux.html) in this way. His preferred definition of austerity is:

>“fiscal consolidation/spending cuts that have a negative impact on aggregate demand which monetary policy is unable to offset.”

So using this definition it’s clear that the fiscal consolidations undertaken by several governments from 2010 onwards would qualify as austerity policies. They were contractionary on demand, and because central banks had already reduced interest rates to effectively zero^[What economists call the “zero lower bound (ZLB)”], the negative impact on aggregate spending could not be offset.

## How much austerity has taken place?
The evidence on austerity measures in the UK will be examined here. The data used is all taken from the UK Office for Budget Responsibility’s [Public finances databank](https://obr.uk/data/). The focus will be on public expenditure rather than taxation revenue as this is where most of the austerity measures have been concentrated. But fiscal consolidation can and does also include increases in taxation. Together, taxation and expenditure changes will determine the overall impact of fiscal consolidation on aggregate demand in the economy - this can be shown by looking at changes in the cyclically adjusted primary balance, something we will do later on. 

Public expenditure comprises two main types: current and capital. Current expenditure is “Spending on items that are 'consumed' in the year of purchase, such as public sector salaries and transfers.” Capital expenditure is spending on new capital assets such as infrastructure (roads, bridges, flood defences etc.), hospital buildings and equipment, school buildings and equipment, other public buildings, and so on. Capital spending is also called investment and can be either gross or net. Gross investment is total spending on capital assets some of which will be replacing capital which has finally worn out (depreciated). Net investment is capital spending which increases the existing capital stock. Thus net investment equals gross investment less depreciation.

### Current Expenditure
Looking first at current spending, figure 1 shows UK public sector **real** current expenditure for the fiscal years starting in 1975-76 up to 2018-19. (As explained in this [post](https://www.bobecon.com/2019/04/how-well-is-the-uk-economy-doing/) it is important to look at real rather than money(nominal) expenditure so we can see whether the volume of goods and services provided by the public sector was increasing or not.) Apart from the occasional, temporary fall (in 77-78, 88-89, and 96-97), there had been a steady increase^[Which was to be expected given the steady growth in real GDP] in real current public expenditure up until 2010, when the “austerity period” began. The impact since 2010-11 is plain to see; real current expenditure has pretty much flatlined. Average, annual (compound) growth in the 20 year period prior to 2010-11 was 3.3%, whilst in the period 2010-11 to 2018-19 it was zero!^[In 2018-19 current real expenditure was £714.8bn, the same as in 2010-11]

```{r echo=FALSE, fig.cap="UK Real Current Expenditure", message=FALSE}
autoplot(UKAust75[,"rcexp"]) + ggtitle("UK Public Sector Current Expenditure\n(£ billion, 2017-18 prices), 1975-76 to 2018-19") + ylab("£ billion") + xlab("") + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + annotate("label", x = 2010, y = 350, label = 'Source: OBR') + scale_x_continuous(limits=c(1975,2020), breaks=seq(1975,2020,by=5))
```

In fact, the impact of austerity has been even greater than figure 1 suggests, as looking at the level of real expenditure alone doesn’t take into account the effect of population growth. A better measure to consider when assessing the impact of the austerity measures is real expenditure *per head of population*. Figure 2 uses the same data as figure 1, but this time these values have been divided by the UK population for each year^[Specifically, these are the mid-year estimates of the UK population (Source: ONS, code UKPOP)]. Again the contrast between the periods before and after 2010-11 is striking. Prior to 2010 the growth in current spending outpaced population growth such that expenditure per capita increased at a consistent rate; since 2010 it has not kept pace and real spending per head has fallen significantly. In 2018-19 real UK public sector current expenditure per head was 5.5% lower than in 2010-11.

```{r echo=FALSE, fig.cap="UK Real Current Expenditure per capita", message=FALSE}
currexppop <- ts(UKAust75[,"rcexp"]/UKAust75[,"ukpop"], start = 1975)
currexppop <- currexppop*1e+09
autoplot(currexppop) + ggtitle("UK Public Sector Current Expenditure\n(£ per capita, 2017-18 prices), 1975-76 to 2018-19") + ylab("£ billion") + xlab("") + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + annotate("label", x = 2010, y = 6000, label = 'Source: OBR') + scale_x_continuous(limits=c(1975,2020), breaks=seq(1975,2020,by=5))
```

### Capital Expenditure
Turning now to UK public sector real capital (investment) expenditure, figures 3 and 4 show real net investment in overall terms and per head of population respectively. Both charts show a similar picture. Net investment by the public sector is much more volatile than current spending with significant fluctuations in the 1980s and 1990s.

```{r echo=FALSE, fig.cap="UK Real Capital Expenditure", message=FALSE}
autoplot(UKAust75[,"rnetinv"]) + ggtitle("UK Public Sector Net Investment\n(£ billion, 2017-18 prices), 1975-76 to 2018-19") + ylab("£ billion") + xlab("") + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + annotate("label", x = 2010, y = 25, label = 'Source: OBR') + scale_x_continuous(limits=c(1975,2020), breaks=seq(1975,2020,by=5))
```

However, the 2000s saw a sustained rise from a low of £218 per capita in 2000-01 to a high of £1000 per person in 2008-09. Then in the period of austerity from 2010-11 to 2013-14, public net investment was cut significantly reaching a low of about £510 per head. It has since risen again, with some fluctuations, back up to £615 per capita in 2018-19, but still well below the level reached before the austerity policy began.^[Remember all these figures are in 2017-18 prices]

```{r echo=FALSE, fig.cap="UK Real Capital Expenditure per capita", message=FALSE}
netinvppop <- ts(UKAust75[,"rnetinv"]/UKAust75[,"ukpop"], start = 1975)
netinvppop <- netinvppop*1e+09
autoplot(netinvppop) + ggtitle("UK Public Sector Net Investment\n(£ per capita, 2017-18 prices), 1975-76 to 2018-19") + ylab("£ billion") + xlab("") + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + annotate("label", x = 2010, y = 250, label = 'Source: OBR') + scale_x_continuous(limits=c(1975,2020), breaks=seq(1975,2020,by=5))
```

### Overall Impact
It's clear, then, that as far as public expenditure is concerned a great deal of cutting back has taken place in the UK since 2010. It has been particularly marked in the case of capital spending, which is often the case.^[It’s easier to cancel or cut back on capital projects rather than cut the wages or numbers of public sector employees] But, recalling the definition of austerity given above, what was the overall impact of these fiscal consolidation measures, which also included some tax rises, on aggregate expenditure and hence GDP? A common method of assessing the total effect of fiscal policy changes is to look at changes in the cyclically-adjusted, primary balance (measured  as a percent of GDP). The primary balance is the difference between the public sector’s income and expenditure, excluding interest payments on the public sector’s debt. “Cyclical-adjusted” means that it is a measure of what the primary balance would be if the economy was at its potential, full employment level of GDP. An increase in the deficit (or fall in the surplus) is an indication of a more expansionary fiscal policy (or expansionary fiscal stance as it’s often called). Conversely, a fall in the deficit (rise in the surplus) indicates a more contractionary fiscal stance.

```{r echo=FALSE, fig.cap="UK Cyclically-adjusted Primary Balance", message=FALSE}
autoplot(UKAust75[,"cycpbalgdp"]) + ggtitle("UK Cyclically Adjusted Primary Balance\n(Percent of GDP), 1975-76 to 2018-19") + ylab("Percent") + xlab("") + theme(plot.title=element_text(size=12, family="Verdana", face="bold", colour="#2A62C0")) + annotate("label", x = 2010, y = 1.0, label = 'Source: OBR') + scale_x_continuous(limits=c(1975,2020), breaks=seq(1975,2020,by=5))
```

Figure 5 makes two things very clear. Firstly, the initial response of the UK government to the financial crisis and the Great Recession that followed was to initiate a significant fiscal expansion aimed at propping up aggregate demand once the conventional monetary policy response of cutting interest rates had reached the ZLB. So, the cyclical-adjusted balance went from -2.0 % of GDP in 2007-08 to -6.3% in 2009-10. Secondly, the move to an equally significant (if not more so) fiscal contraction (“the turn to austerity”) from 2010-11 onwards. Apart from a couple of temporary reversals in 2012-13 and 2014-15, the cyclically-adjusted primary deficit has been reduced from -6.3% in 2009-10 to only -0.4% in 2018-19.
